---
title: "Arkesh Das CMSE 410 Semester Project"
author: "Arkesh Das"
date: "03/30/2025"
output:
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
    includes:
      in_header: preamble.tex
fontsize: 12pt
mainfont: Times New Roman
monofont: Courier New
linestretch: 2
---
# Assessing the Influence of False Discovery Rate Methods on Genetic Associations in an Immune Response GWAS

## Background 

Genome-wide association studies (GWAS) have emerged as powerful tools for linking genetic variants with complex traits, including immune responses. A typical GWAS involves testing millions of single nucleotide polymorphisms (SNPs) for associations, which leads to a substantial multiple testing burden. Applying a simple p-value threshold (e.g., 0.05) is inadequate, as it would generate many false positives. Thus, false discovery rate (FDR) control has become an essential strategy to address this challenge.


## About the data

This project utilizes data from the Milieu Intérieur (MI) project, a comprehensive, population-based study coordinated by Prof. Lluis Quintana-Murci and Dr. Darragh Duffy at Institut Pasteur in Paris. Established under the French Government's Investissement d'Avenir – Laboratoire d’Excellence (LabEx) initiative, the MI project aims to dissect the interplay between genetics, environment, and immune variation [1]. 

It was designed to address a critical discrepancy in medicine: while immune responses vary widely among individuals, medical care and therapeutic strategies are often standardized across populations.

 I chose to use this dataset because it offers a unique opportunity to evaluate the impact of false discovery rate methods on GWAS results in the context of immune variation, and the data is high-resolution and well-characterized so it’s ideal for benchmarking statistical methods.

The data was accessed from the NHGRI-EBI GWAS Catalog. The Catalog is a publicly available repository of SNP-trait associations identified in published GWAS studies. It provides standardized annotations, p-values, effect sizes, and sample information for millions of SNPs across hundreds of phenotypes. 

The MI cohort consists of 1,000 healthy individuals, age- and sex-stratified across five decades (ages 20-70), and of French origin. Detailed serological data were collected, including total levels of IgA, IgE, IgG, and IgM, along with antibody responses to 15 antigens from common infectious agents and vaccines (e.g., influenza, EBV, CMV, HSV, rubella). The cohort was genotyped for over 700,000 SNPs and subsequently imputed to yield over 12 million genetic variants. Extensive phenotypic, demographic, clinical, and environmental metadata are also available, including vaccine history, infection exposure, CRP levels, lipid profiles, and lifestyle factors [2].

# Loading in the Data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knitr::opts_chunk$set(fig.width=12, fig.height=8) 

library(tidyverse)
library(qvalue)

theme_gwas <- function(base_size = 12) {
  theme_classic(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 2),
      plot.subtitle = element_text(color = "gray30"),
      axis.title = element_text(face = "bold"),
      axis.text.x = element_text(size = base_size - 2),
      panel.grid.major.y = element_line(color = "gray88", linewidth = 0.25),
      legend.position = "none"
    )
}
```

# Reading in the Data and making dataframes

```{r dataframes, include=FALSE}

#df will contain all of the data from the study that had MAF > 5%
#df2 is 1,000 snps, used to test code chunks before running on larger data frames
#df3 contains 100,000 random snps, used as an analogue for the entire data set
#ch6 contains the 354095 snps identified on chromosome 6

gwas_cols <- c("#CHROM", "POS", "ID", "REF", "ALT", "ALT_FREQ",
               "TEST", "OBS_CT", "OR", "SE", "T_STAT", "P")

data_path <- Sys.getenv("GWAS_DATA_PATH", unset = "HBV_HBc_GWAS_serostatus.txt")

if (!file.exists(data_path)) {
  stop(
    "Could not find the GWAS input file. Place HBV_HBc_GWAS_serostatus.txt ",
    "in the project root, or set GWAS_DATA_PATH to its location."
  )
}

# Reloading the full dataframe takes a long time because of the number of observations.
df <- read.table(data_path, header = TRUE, sep = "", stringsAsFactors = FALSE,
                 check.names = FALSE, comment.char = "")

missing_cols <- setdiff(gwas_cols, names(df))
if (length(missing_cols) > 0) {
  stop("Missing expected columns: ", paste(missing_cols, collapse = ", "))
}

df <- df[, gwas_cols]
head(df)

df <- df %>%
  mutate(
    `#CHROM` = as.integer(`#CHROM`),
    POS = as.numeric(POS),
    P = as.numeric(P)
  )

df_clean <- df %>%
  filter(!is.na(P), P > 0, P <= 1)

data_summary <- df_clean %>%
  summarise(
    valid_p_values = n(),
    min_p_value = min(P),
    bonferroni_threshold = 0.05 / n()
  )

data_summary

df2 <- head(df_clean, 1000)
head(df2)

set.seed(03302025)  # Setting seed for reproducibility for df3
sample_size <- min(100000, nrow(df_clean))
df3 <- df_clean[sample(nrow(df_clean), sample_size), ]
colnames(df3) <- gwas_cols
head(df3)

ch6 <- df_clean %>% filter(`#CHROM` == 6)
colnames(ch6) <- gwas_cols
head(ch6)

```

After filtering out missing or invalid p-values, the dataset contains `r data_summary$valid_p_values` valid tests. The smallest raw p-value is `r signif(data_summary$min_p_value, 3)`, while the Bonferroni threshold for the full filtered dataset is `r signif(data_summary$bonferroni_threshold, 3)`. This means that even the strongest raw association in this file does not pass a full-data Bonferroni correction.

# Creating Manhattan Plots

Because I will be making several Manhattan, QQ, and adjusted p-value plots, I created reusable plotting functions. This keeps the visual style consistent and makes it easier to compare results across FDR methods.

```{r manhattan plot function, include=FALSE}
prepare_manhattan_data <- function(df,
                                   chrom_col = "#CHROM",
                                   pos_col = "POS",
                                   pval_col = "P") {
  df_plot <- df %>%
    mutate(
      CHR = as.integer(gsub("[^0-9]", "", as.character(.data[[chrom_col]]))),
      BP = as.numeric(.data[[pos_col]]),
      p_value = as.numeric(.data[[pval_col]])
    ) %>%
    filter(!is.na(CHR), !is.na(BP), !is.na(p_value), p_value > 0, p_value <= 1) %>%
    arrange(CHR, BP)

  chr_offsets <- df_plot %>%
    group_by(CHR) %>%
    summarise(max_bp = max(BP), .groups = "drop") %>%
    arrange(CHR) %>%
    mutate(offset = lag(cumsum(max_bp), default = 0))

  df_plot %>%
    left_join(chr_offsets, by = "CHR") %>%
    mutate(BP_cum = BP + offset)
}

get_axis_df <- function(df_plot) {
  df_plot %>%
    group_by(CHR) %>%
    summarise(center = (min(BP_cum) + max(BP_cum)) / 2, .groups = "drop")
}

plot_manhattan <- function(df,
                           chrom_col = "#CHROM",
                           pos_col = "POS",
                           pval_col = "P",
                           plot_title = "Manhattan Plot",
                           plot_subtitle = NULL,
                           threshold = NULL,
                           y_limit = NULL,
                           highlight = NULL) {
  df_plot <- prepare_manhattan_data(df, chrom_col, pos_col, pval_col)
  axis_df <- get_axis_df(df_plot)

  if (!is.null(highlight) && "ID" %in% names(df_plot)) {
    df_plot <- df_plot %>% mutate(is_highlight = ID %in% highlight)
  } else {
    df_plot <- df_plot %>% mutate(is_highlight = FALSE)
  }

  p <- ggplot(df_plot, aes(x = BP_cum, y = -log10(p_value))) +
    geom_point(aes(color = factor(CHR %% 2)), alpha = 0.65, size = 0.8) +
    geom_point(data = filter(df_plot, is_highlight),
               color = "#B23A48", size = 1.7, alpha = 0.95) +
    scale_color_manual(values = c("#4A4A4A", "#2F80B7")) +
    scale_x_continuous(breaks = axis_df$center, labels = axis_df$CHR,
                       expand = expansion(mult = c(0.01, 0.01))) +
    labs(
      title = plot_title,
      subtitle = plot_subtitle,
      x = "Chromosome",
      y = expression(-log[10](p-value))
    ) +
    theme_gwas()

  if (!is.null(threshold)) {
    p <- p +
      geom_hline(yintercept = -log10(threshold),
                 color = "#B23A48", linetype = "dashed", linewidth = 0.5)
  }

  if (!is.null(y_limit)) {
    p <- p + coord_cartesian(ylim = c(0, y_limit))
  }

  p
}

plot_adjusted_pvalues <- function(df,
                                  pval_col,
                                  plot_title,
                                  plot_subtitle = NULL,
                                  alpha = 0.05) {
  df_plot <- prepare_manhattan_data(df, pval_col = pval_col)
  axis_df <- get_axis_df(df_plot)

  ggplot(df_plot, aes(x = BP_cum, y = p_value)) +
    geom_point(aes(color = factor(CHR %% 2)), alpha = 0.6, size = 0.75) +
    geom_hline(yintercept = alpha, color = "#B23A48",
               linetype = "dashed", linewidth = 0.5) +
    scale_color_manual(values = c("#4A4A4A", "#2F80B7")) +
    scale_x_continuous(breaks = axis_df$center, labels = axis_df$CHR,
                       expand = expansion(mult = c(0.01, 0.01))) +
    coord_cartesian(ylim = c(0, 1)) +
    labs(
      title = plot_title,
      subtitle = plot_subtitle,
      x = "Chromosome",
      y = "Adjusted p-value"
    ) +
    theme_gwas()
}

plot_qq <- function(p_values, plot_title, plot_subtitle = NULL) {
  p_values <- p_values[!is.na(p_values) & p_values > 0 & p_values <= 1]
  n <- length(p_values)
  qq_df <- tibble(
    expected = -log10(ppoints(n)),
    observed = -log10(sort(p_values))
  )

  ggplot(qq_df, aes(x = expected, y = observed)) +
    geom_point(color = "#2F80B7", alpha = 0.55, size = 0.9) +
    geom_abline(slope = 1, intercept = 0, color = "#B23A48",
                linewidth = 0.55, linetype = "dashed") +
    coord_equal() +
    labs(
      title = plot_title,
      subtitle = plot_subtitle,
      x = expression(Expected ~ -log[10](p-value)),
      y = expression(Observed ~ -log[10](p-value))
    ) +
    theme_gwas()
}

```

```{r testing function, echo = FALSE, message = FALSE, warning = FALSE}
plot_manhattan(
  df3,
  threshold = 0.05,
  plot_title = "Raw GWAS associations in the 100,000-SNP sample",
  plot_subtitle = "Dashed line marks the uncorrected p = 0.05 threshold"
)
```

The first plot uses the raw p-values from the 100,000-SNP sample and overlays a Bonferroni threshold. This is a useful baseline because it shows how stringent family-wise error control becomes in a GWAS setting.

```{r raw 100k bonferroni, echo = FALSE, message = FALSE, warning = FALSE}
df3_clean <- df3 %>%
  filter(!is.na(P))
bonferroni_threshold <- 0.05 / nrow(df3_clean)

plot_manhattan(
  df3_clean,
  threshold = bonferroni_threshold,
  y_limit = 10,
  plot_title = "Raw GWAS associations with Bonferroni threshold",
  plot_subtitle = paste0("Threshold: 0.05 / ", nrow(df3_clean), " tests")
)

```

```{r chromosome 6 raw pvalues, echo = FALSE, message = FALSE, warning = FALSE}

ch6_clean <- ch6 %>%
  filter(!is.na(P))
plot_manhattan(
  ch6_clean,
  threshold = 0.05,
  y_limit = 6,
  highlight = "rs4254998",
  plot_title = "Raw GWAS associations on chromosome 6",
  plot_subtitle = "rs4254998 is highlighted; dashed line marks p = 0.05"
)
```
# QQ plots

QQ plots are useful here because they compare the observed p-value distribution with what would be expected under the global null hypothesis. Large departures from the diagonal can indicate true association signal, inflation, or population structure that still needs to be checked.

```{r QQ plot for subset}
plot_qq(
  df3$P,
  plot_title = "QQ plot of raw GWAS p-values",
  plot_subtitle = "100,000-SNP sample"
)
```

```{r qqplot 100k, echo = FALSE, message = FALSE, warning = FALSE}
plot_qq(
  df3_clean$P,
  plot_title = "QQ plot of raw GWAS p-values",
  plot_subtitle = "After filtering missing or invalid p-values"
)
```

```{r qqplot chr6, echo = FALSE, message = FALSE, warning = FALSE}
plot_qq(
  ch6_clean$P,
  plot_title = "QQ plot of chromosome 6 p-values",
  plot_subtitle = "Chromosome-specific subset"
)
```

# Various FDR methods

Now I will be trying different FDR methods, from most to least conservative.

## Benjamini-Yekutieli

Out of the p-value adjustment methods, Benjamini Yekuteli is the most conservative, so I will try it first. R has a `p.adjust` function that can automatically apply the BY method.

```{r Benjamini-Yekutieli test, echo = FALSE, message = FALSE, warning = FALSE}

df3_clean$P_BY <- p.adjust(df3_clean$P, method = "BY")

# Summary stats of BY-adjusted p-values
summary(df3_clean$P_BY)

# Filter significant results under BY 
df3_by_sig <- df3_clean %>% filter(P_BY < 0.05)

plot_adjusted_pvalues(
  df3_clean,
  pval_col = "P_BY",
  plot_title = "BY-adjusted p-values across the genome",
  plot_subtitle = "Dashed line marks FDR = 0.05"
)
```

Because BY-adjusted p-values are adjusted significance values rather than raw association p-values, I plot them on the adjusted p-value scale instead of forcing them into a standard Manhattan plot.

```{r BY plot test, echo = FALSE, message = FALSE, warning = FALSE}
plot_adjusted_pvalues(
  df3_clean,
  pval_col = "P_BY",
  plot_title = "BY-adjusted p-values across the genome",
  plot_subtitle = "Most adjusted values are near 1, reflecting the conservatism of BY"
)
```

Using this adjusted p-value plot, we can see that the Benjamini-Yekutieli method is highly conservative for this dataset. In the 100,000-SNP sample, it does not identify any variants below an FDR threshold of 0.05.

## Benjamini-Hochberg

Now I will try adjust my p-values using the Benjamini-Hochberg method.

```{r Benjamini-Hochberg test, echo = FALSE, message = FALSE, warning = FALSE}
# Adjusting p_values using BH
df3_clean$P_BH <- p.adjust(df3_clean$P, method = "BH")

#Summary stats of adjusted p-values
summary(df3_clean$P_BH)

df3_bh_sig <- df3_clean %>% filter(P_BH < 0.05)
df3_clean$log10_P_BH <- -log10(df3_clean$P_BH)

plot_adjusted_pvalues(
  df3_clean,
  pval_col = "P_BH",
  plot_title = "BH-adjusted p-values across the genome",
  plot_subtitle = "Dashed line marks FDR = 0.05"
)


```


```{r BH plot test, echo = FALSE, message = FALSE, warning = FALSE}
plot_adjusted_pvalues(
  df3_clean,
  pval_col = "P_BH",
  plot_title = "BH-adjusted p-values across the genome",
  plot_subtitle = "BH is less conservative than BY, but still controls FDR"
)
```

## Benjamini-Krieger-Yekutieli 

Now I will try to use the Benjamini-Krieger-Yekutieli method in the `mutoss` package [4]: 
```{r mutoss BKY, eval = FALSE}
mutoss::loadMethod("BKY")


bky_result <- mutoss::runMethod("BKY", list(pValues = df3_clean$P, alpha = 0.05))
```

Unfortunately, I could not get the `mutoss` package's implementation of BKY to work, so I will instead follow the original 2006 paper [5], which describes how to implement the BKY method:

```{r BKY implementation}
# BKY two-stage adaptive FDR adjustment.
bky_adjust <- function(p, alpha = 0.05) {
  valid <- !is.na(p)
  adjusted <- rep(NA_real_, length(p))
  p_valid <- p[valid]
  m <- length(p_valid)

  if (m == 0) {
    return(adjusted)
  }

  stage_one_alpha <- alpha / (1 + alpha)
  stage_one_reject <- p.adjust(p_valid, method = "BH") <= stage_one_alpha
  r1 <- sum(stage_one_reject)

  if (r1 == 0) {
    adjusted[valid] <- 1
    return(adjusted)
  }

  m0_hat <- max(1, m - r1)
  adjusted[valid] <- pmin(1, p.adjust(p_valid, method = "BH") * (m0_hat / m))
  adjusted
}

# Applying BKY to the data

df3_clean$P_BKY <- bky_adjust(df3_clean$P, alpha = 0.05)
df3_bky_sig <- df3_clean %>% filter(P_BKY < 0.05)

summary(df3_clean$P_BKY)
nrow(df3_bky_sig)

plot_adjusted_pvalues(
  df3_clean,
  pval_col = "P_BKY",
  plot_title = "BKY-adjusted p-values across the genome",
  plot_subtitle = "Two-stage adaptive FDR adjustment"
)

```

```{r BKY plot test, echo = FALSE, message = FALSE, warning = FALSE}
plot_adjusted_pvalues(
  df3_clean,
  pval_col = "P_BKY",
  plot_title = "BKY-adjusted p-values across the genome",
  plot_subtitle = "Dashed line marks FDR = 0.05"
)
```
Because BKY estimates the number of true null hypotheses before applying its second-stage threshold, it can be slightly less conservative than the standard BH procedure. In this sample, the important comparison is the number of discoveries, not just whether the adjusted p-values look visually separated on the plot.

## Storey-Tibshirani (q-value)

Now I will be adjusting my p-values using the Storey-Tibshirani method [6]. Thankfully, there is a package in R that was designed in collaboration with the creators of this method, the `qvalue` package. I followed the documentation of the `qvalue` package to use the `qvalue()` method and other methods on my data [7].

```{r, echo = FALSE, message = FALSE, warning = FALSE}
qobj <- qvalue(p = df3_clean$P)  # estimating pi_0 and q-values
df3_clean$q_value <- qobj$qvalues

df3_q_sig <- df3_clean %>% filter(q_value < 0.05)
nrow(df3_q_sig)  

qobj$pi0

```

The Storey-Tibshirani q-value method estimates the proportion of true null hypotheses before calculating q-values. In my run, the estimated pi_0 was 0.971, meaning that about 97.1% of the tested SNPs were estimated to be null.

```{r FDR summary table, echo = FALSE, message = FALSE, warning = FALSE}
fdr_summary <- tibble(
  method = c("Benjamini-Yekutieli", "Benjamini-Hochberg",
             "Benjamini-Krieger-Yekutieli", "Storey-Tibshirani q-value"),
  significant_variants = c(
    nrow(df3_by_sig),
    nrow(df3_bh_sig),
    nrow(df3_bky_sig),
    nrow(df3_q_sig)
  )
)

knitr::kable(fdr_summary, col.names = c("Method", "Variants below 0.05"))
```

Overall, the different FDR approaches tell a consistent story: after correcting for multiple testing, the 100,000-SNP sample does not provide strong evidence for a large number of robust associations. This does not mean that the original study has no signal. It means that a random subset of the full GWAS is best used here as a method comparison, while definitive association claims should be made from the full dataset.

```{r best BKYs, echo = FALSE, message = FALSE, warning = FALSE}
top_bky <- df3_clean %>%
  arrange(P) %>%
  slice(1:50) %>%
  select(`#CHROM`, POS, ID, P, P_BKY)

top_chr6_candidate <- ch6_clean %>%
  arrange(P) %>%
  slice(1) %>%
  select(`#CHROM`, POS, ID, P)

top_chr6_candidate
```

Taking a second look at the smallest raw p-values, there is at least one chromosome 6 candidate worth inspecting because chromosome 6 contains the HLA region, which was important in the original study. I would treat this as an exploratory follow-up rather than a confirmed discovery from this project alone.

Top chromosome 6 candidate in the full chromosome 6 subset:

- Chromosome: `r top_chr6_candidate[["#CHROM"]][1]`
- Position: `r top_chr6_candidate$POS[1]`
- SNP ID: `r top_chr6_candidate$ID[1]`
- Raw p-value: `r signif(top_chr6_candidate$P[1], 3)`

I used the NCBI Human Genome Data Viewer [8] as a first-pass way to inspect candidate genomic context. The biological interpretation should remain conservative unless the genome build, SNP ID, and nearest-gene annotation are confirmed from the same reference build used in the GWAS summary statistics. If the top chromosome 6 signal maps near immune-related loci, it would be a reasonable candidate for follow-up, but the evidence here is not strong enough to claim a novel immune-response locus.

Therefore, my conclusion is that the FDR method choice strongly changes the apparent number of discoveries, especially when moving from raw p-values to multiple-testing-adjusted results. BY and BH are conservative in this sample, Storey's q-value approach estimates that most tests are null, and BKY provides an adaptive comparison point. The strongest portfolio-level result is the statistical comparison itself: this project shows why multiple-testing correction is necessary in GWAS and why candidate SNP interpretation has to be tied to corrected significance, genome build, and biological annotation.


#Misc graphs


```{r normal distribution testing, echo = FALSE, message = FALSE, warning = FALSE}
df3_clean$logP <- -log10(df3_clean$P)


ggplot(df3_clean, aes(x = -log10(P))) +
  geom_histogram(bins = 50, fill = "#2F80B7", color = "white") +
  coord_cartesian(xlim = c(0, 4)) +
  labs(title = "Distribution of -log10(p-values)",
       x = expression(-log[10](p-value)), y = "SNP count") +
  theme_gwas()


ggplot(df3_clean, aes(x = -log10(P))) +
  geom_histogram(bins = 50, fill = "#2F80B7", color = "white") +
  coord_cartesian(xlim = c(2, 4)) +
  labs(title = "Right tail of -log10(p-values)",
       x = expression(-log[10](p-value)), y = "SNP count") +
  theme_gwas()

```

```{r normal dist curve, echo = FALSE, message = FALSE, warning = FALSE}
#To create figure in slide deck
x <- seq(-4, 4, length.out = 1000)
y <- dnorm(x)

df_norm <- data.frame(x = x, y = y)

z_cutoff <- qnorm(0.95)  

shade_df <- subset(df_norm, x > z_cutoff)


ggplot(df_norm, aes(x, y)) +
  geom_line(color = "black", linewidth = 0.8) +
  geom_area(data = shade_df, aes(x, y), fill = "#2F80B7", alpha = 0.55) +
  geom_vline(xintercept = z_cutoff, color = "#B23A48",
             linetype = "dashed", linewidth = 0.6) +
  labs(
    title = "Standard Normal Distribution with One-Sided p = 0.05 Cutoff",
    x = "Z-score",
    y = "Density"
  ) +
  theme_gwas()

```

# References

[1] https://www.milieuinterieur.fr/en/about-us/the-milieu-interieur/

[2] Scepanovic, P., Alanio, C., Hammer, C., et al. (2018). Human genetic variants and age are the strongest predictors of humoral immune responses to common pathogens and vaccines. Genome Medicine, 10, 59. https://doi.org/10.1186/s13073-018-0568-8

[3] Wickham, H. (2016). ggplot2: Elegant Graphics for Data Analysis. Springer-Verlag New York. https://ggplot2.tidyverse.org/

[4] https://search.r-project.org/CRAN/refmans/mutoss/html/multiple.down.html

[5] Benjamini, Y., Krieger, A. M., & Yekutieli, D. (2006). Adaptive Linear Step-up Procedures That Control the False Discovery Rate. Biometrika, 93(3), 491–507. https://doi.org/10.1093/biomet/93.3.491 

[6] Storey JD, Tibshirani R. Statistical significance for genomewide studies. Proc Natl Acad Sci U S A. 2003 Aug 5;100(16):9440-5. doi: 10.1073/pnas.1530509100. Epub 2003 Jul 25. PMID: 12883005; PMCID: PMC170937.

[7] https://www.rdocumentation.org/packages/qvalue/versions/2.4.2/topics/qvalue

[8] https://www.ncbi.nlm.nih.gov/gdv/
